# Static digital images

The three fundamental quantities of sound waves are time, frequency, and amplitude. The most commonly used visualisations of sounds are consequently combinations of these properties.

The `seewave` package [@seewave] is the easiest way to produce these plots. We will use the `sheep` audio sample provided by `seewave` for the first plots.

```{r}
library(seewave)
data(sheep, package="seewave")
```

## The time domain

_Time domain_ techniques analyse the recorded waveform directly. Conversion of a waveform to a spectrum of frequencies is a relatively costly computation. Before digital computers, frequency analysis was hard to achieve with the precision and resolution we are accustomed to today. 

### Oscillograms

Digitally sampled audio is a series of amplitude measurements sampled at a regular time interval. The most basic plot we could recreate from a sampled file would be plotting amplitude values against time, a plot known as an oscillogram.

```{r, amplitude-bars,  fig.width=9, fig.cap='Plotting a waveform', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
barplot(sheep@left[8001:8050])
```

Figure \@ref(fig:amplitude-bars) shows 50 samples from the sheep audio file plotted using the standard R barplot()function. This representation has the advantage that the individual bars are a visual reminder that we are handling a sampled representation instead of a continuously varying waveform. It is common, however, to create a visual representation closer to that of an actual wave (Figure \@ref(fig:amplitude-lines) shows how this is achieved using base R functions).

```{r, amplitude-lines,  fig.width=9, fig.cap='Plotting a waveform', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
plot(sheep@left[8001:8050], type="l")
```

The `oscillo()` function from `seewave` handles the labelling of axes for us and is a very convenient solution. Figure \@ref(fig:amplitude-oscillo) shows the resulting plot, an oscillogram, for the entire `sheep` file.

```{r, amplitude-oscillo,  fig.width=9, fig.cap='Oscillogram', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
oscillo(sheep)
```

## The frequency domain

The _frequency domain_ covers analyses in which frequency data are extracted from the sampled waveform, typically using the _Fast Fourier Transform_ algorithm.

### Plotting a spectrum

A spectrum is a plot of amplitude against frequency.

```{r, spectrum,  fig.width=9, fig.cap='Spectrum', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
spec(sheep)
```

### Plotting a spectrogram


```{r, spectrogram,  fig.width=9, fig.cap='Spectrogram', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
spectro(sheep)
```

#### Settings for a FFT