# Patterns of activity
Lots of organismal activities are tied to the cycles of the day, and particularly in temperate zones, cycles of the year. These cycles bring regular fluctuations in light levels, day lengths, temperatures, and a host of other influences. Often these cycles interact, with the dawn chorus peaking in the early daylight hours, and it's timing and intensity fluctuating on a yearly cycle. This chapter looks at visualising these cycles, and additionally the effects of lunar cycles.

These plots are created using the SonicScrewdriver package [@sonicscrewdriver] which in turn uses the suncalc package [@suncalc] to perform the required sun and moon position calculations. The Plotrix package [@plotrix] is used for creating the visualisation. These packages can be installed as shown below.

```{r eval=FALSE}
install.packages(c("plotrix", "sonicscrewdriver"))
```

The SonicScrewdriver package must be loaded before constructing a visual.

```{r message=FALSE}
library(sonicscrewdriver)
```

## Daily Cycles { #daily-cycles }
The use of the term _diel_ for daily cycles has been contested by @broughton1963 as being an incorrectly formed unnecessary neologism, it sees greater use (according to the online Oxford English Dictionary) than his suggested _nycthemeral_.

The design for these plots came from a desire to compare the dawn chorus at various locations around the UK, although they also offer great potential for comparing locations with greater longitudinal and/or latitudinal separation. The plots show the times of day, night, twilight (\@ref(twilight-types)), sunrise, sunset, nadir and solar noon. The day part of the plot shows the altitude (angle of the sun above the horizon) throughout the day, with the maximum value representing the sun being directly overhead.

### The Types of Twilight { #twilight-types }

```{r, twilights, fig.asp=0.6875, fig.width=9, fig.cap='Different types of twilight as the sun sets below the horizon', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
plot.default(
  NULL, 
  NULL, 
  xlim=c(-0.2,1.4),
  ylim=c(-1,0.1),
  axes=FALSE,
  xlab="",
  ylab=""
)

# Horizon
lines(c(0,1), c(0,0))

text(0,0,"horizon",cex=0.8,adj=c(1,0))

degs <- c(18,12,6)
rads <- degs*2*pi/360
labs <- c("astronomical", "nautical", "civil")
fills <- c(rgb(0.4,0.4,0.4,1), rgb(0.6,0.6,0.6,1), rgb(0.8,0.8,0.8,1))
labs <- paste0("-",degs,"°: ",labs)

sunset_y <- -tan(rads)

for (i in 1:length(sunset_y)) {
  x <- c(0,1,1)
  y <- c(0,0,sunset_y[i])
  polygon(x,y, col=fills[i])
  text(1.01,sunset_y[i], labs[i], adj=c(0,0), cex=0.8)
}
```

#### Civil Twilight
Civil twilight occurs when the geometric centre of the sun (as seen from Earth) passes between 0° and 6° below the horizon. During this time it is normal for humans not to need the assistance of artifical light for everyday tasks.

#### Nautical Twilight
Nautical twilight occurs when the sun is between 6° and 12° below the horizon. During this time there is sufficient light to distinguish the horizon even without illumination from the moon (allowing determination of position at sea through star sightings).

#### Astronomical Twilight
When the sun is between 12° and 18° below the horizon many astronomical observations are possible even though some light from the sun is visible through the atmosphere. In urban areas with light pollution this is often considered to be a dark sky.

### Diel Plots {#diel-plots }
As the times of the solar day are dependent both on the date and location these must be passed to the `dielPlot()` function.


```{r diel-plot-1, fig.height=7, fig.width=9, fig.cap='Example of a diel plot', fig.align='center', out.width='90%', cache=TRUE}
dielPlot("2022-08-08", lat=53, lon=0.1)
```

#### Rotations of a `dielPlot()`

By default the information is plotted in the UTC timezone, so locations in other timezones will have an overall rotation.

```{r diel-plot-no-tz, fig.height=7, fig.width=9, fig.cap='Diel plots in UTC.', fig.align='center', out.width='90%', cache=TRUE}
par(mfrow=c(1,3))
dielPlot(Sys.Date(), lat=53, lon=-50)
dielPlot(Sys.Date(), lat=53, lon=-0)
dielPlot(Sys.Date(), lat=53, lon=50)
```

Plots can be made in any timezone by using the `rot` parameter to `dielPlot()` and the `tz()` function.

```{r diel-plot-tz, fig.height=7, fig.width=9, fig.cap='The same diel plots in UTC +3', fig.align='center', out.width='90%', cache=TRUE}
par(mfrow=c(1,3))
dielPlot(Sys.Date(), lat=53, lon=-50, rot=tz(3))
dielPlot(Sys.Date(), lat=53, lon=-0, rot=tz(3))
dielPlot(Sys.Date(), lat=53, lon=50, rot=tz(3))
```

By setting the `rot` parameter to `Solar Noon` it is possible to align the plots to solar noon. Notice that this rotates the plot labels.

```{r diel-plot-sn, fig.height=7, fig.width=9, fig.cap='The same diel plots aligned to solar noon', fig.align='center', out.width='90%', cache=TRUE}
par(mfrow=c(1,3))
dielPlot(Sys.Date(), lat=53, lon=-50, rot="Solar Noon")
dielPlot(Sys.Date(), lat=53, lon=-0, rot="Solar Noon")
dielPlot(Sys.Date(), lat=53, lon=50, rot="Solar Noon")
```

#### Customising a `dielPlot()`

In addition to the `date`, `lat` and `lon` parameters to `dielPlot()` it is possible to make additional customisations to how the information is presented.

__Legend__

A legend can be added to the plot by setting `legend=TRUE`.

```{r diel-plot-legend, fig.height=7, fig.width=9, fig.cap='Example adding a legend to a diel plot', fig.align='center', out.width='90%', cache=TRUE}
dielPlot("2022-08-08", lat=53, lon=0.1, legend=TRUE)
```


__Plotting Components__

The components that can be plotted are listed below. By default all are plotted except for `Solar Noon` and `Nadir`. 

Name |Notes
-----|------
Astronomical Twilight |
Nautical Twilight |
Civil Twilight |
Sunrise |
Solar Noon | The time when the sun is highest in the sky
Sunset |
Nadir |

The components that are plotted can be specified using the `plot` parameter.

```{r diel-plot-components, fig.height=7, fig.width=9, fig.cap='Selecting the components to plot', fig.align='center', out.width='90%', cache=TRUE}
components <- c("Sunrise", "Sunset", "Solar Noon", "Nadir")
dielPlot("2022-08-08", lat=53, lon=0.1, plot=components)
```

## Yearly Cycles

The `yearlyPlot()` function from the SonicScrewdriver package shows daylight changes throughout a year. It behaves in a very similar fashion to `dielPlot` but takes a single year rather a date as input.

```{r yearly-plot-1, fig.height=7, fig.width=9, fig.cap='Example of a yearly plot', fig.align='center', out.width='90%', cache=TRUE}
yearlyPlot(2022, lat=53, lon=0.1)
```

## Lunar Cycles

## Core and ring plots

These visualisations for cyclical data plot their information onto a circle with radius of two units. It is possible to limit the plot either to the centre of the circle (a 'core' plot) or to the edge (a 'ring plot'). These alternative forms may be more useful when these plots are used to visualise addition variables (\@ref(adding-to-cyclical)).

```{r diel-plot-core, fig.height=7, fig.width=9, fig.cap='A \'core\' diel plot.', fig.align='center', out.width='90%', cache=TRUE}
dielPlot("2022-08-08", lat=53, lon=0.1, limits=c(0,1))
```

```{r diel-plot-ring, fig.height=7, fig.width=9, fig.cap='A \'ring\' diel plot.', fig.align='center', out.width='90%', cache=TRUE}
dielPlot("2022-08-08", lat=53, lon=0.1, limits=c(1,2))
```

## Behind the scenes

### `radialPolygon()`

The majority of the plotting performed for cyclical plots in `SonicScrewdriver` is performed by the `radialPolygon()` function. This function can be used to plot sectors, annuli, horizon plots and irregular polygons. It is used by the plotting functions such as `dielPlot()` as well as helper functions such as `dielRings()` to add data to cyclical plots.

For simple use cases, knowledge of the operation of `radialPolygon()` may not be needed. A number of helper functions cover the most common uses. However, an understanding of how this function works will allow for far greater customisation of cyclical plots than would otherwise be possible.

```{r radialPolygon-types, fig.height=7, fig.width=9, fig.cap='Types of radial polygon plot', fig.align='center', out.width='90%', echo=FALSE, cache=TRUE}
emptyDiel()

radialPolygon(pi/6,pi/3, 0, 2,col="red")
radialPolygon(pi/2,pi,1.75,2, col="blue")

wave <- c(0,0.25,0.5,0.25,0,0.25,0.5,0.25)
angs <- pi + 1:8 * pi/16
radialPolygon(NA,angs,1,1.25+wave, col="orange")
angs <- angs+ pi/2
radialPolygon(angs,angs,1+rev(wave), 1.25+wave, col="green")

legend(
  -3,2.5,
  c("sector", "annulus", "horizon", "polygon"),
  col=c("red", "blue", "orange", "green"),
  lty=1,
  lwd=5,
  bty = "n",
  cex = 1)
```

The various types of plots are created by changing the angle and radius parameters to `radialPolygon().`

```{r eval=FALSE}
radialPolygon(angle1, angle2, radius1, radius2, col)
```

#### Orientation

Unlike traditional polar plots, diel and yearly plots start their periods on the left hand horizontal, and proceed clockwise. This orientation is assumed by `radialPolygon()`, although it may be modified (e.g. the parameters `reverse=FALSE` and `rot=0` will plot using the standard conventions for polar coordinate systems.)

#### Sectors

A sector is a section of a circle defined by two radii and an arc between them. Sectors are used widely in the default settings of `dielPlot()` to plot the times of night and twilight.

```{r radialPolygon-sector, fig.height=7, fig.width=9, fig.cap='Plotting a sector', fig.align='center', out.width='90%', cache=TRUE}
emptyDiel()
radialPolygon(pi/6, 2*pi/3, 0, 2, col="red")
```

Reversing the angle arguments allows the complementary sector to be drawn.

```{r radialPolygon-sector2, fig.height=7, fig.width=9, fig.cap='Plotting a sector', fig.align='center', out.width='90%', cache=TRUE}
emptyDiel()
radialPolygon(2*pi/3, pi/6, 0, 2, col="blue")
```

#### Annuli

An annulus is the region between two concentric circles. Annuli and annular sectors are generated by `radialPolygon()` when the parameter `radius` is greater than zero.

```{r radialPolygon-annuli, fig.height=7, fig.width=9, fig.cap='Plotting an annulus and an annular sector', fig.align='center', out.width='90%', cache=TRUE}
emptyDiel()
radialPolygon(0, 2*pi, 1.75, 2, col="blue")
radialPolygon(pi, 4*pi/3, 1, 1.5, col="red")
legend(
  -3,2.5,
  c("annulus", "annuluar sector"),
  col=c("blue", "red"),
  lty=1,
  lwd=5,
  bty = "n",
  cex = 1)
```

#### Horizons

Horizons have one circular edge, and one that represents data, they are named as they often resemble a landscape or cityscape horizon. The example below uses a generated sine pattern to form the data edge.

```{r radialPolygon-horizon, fig.height=7, fig.width=9, fig.cap='Plotting a horizon polygon.', fig.align='center', out.width='90%', cache=TRUE}
library(tuneR)

angles <- (0:200)*pi/200 + pi/2
values <- 0.05*sine(10, samp.rate=201)@left

emptyDiel()
radialPolygon(NA,angles,0.5,1+values)
```

Setting the first angle parameter to `NA` uses the range of the second to calculate the inner edge.

The inner edge can be used to show data by swapping the order of the angle and radius parameters.

```{r radialPolygon-horizon2, fig.height=7, fig.width=9, fig.cap='Plotting a horizon polygon.', fig.align='center', out.width='90%', cache=TRUE}
library(tuneR)

angles <- (0:200)*pi/200 + pi/2
values <- 0.05*sine(10, samp.rate=201)@left

emptyDiel()
radialPolygon(angles,NA,1+values,2)
```

The `yearlyPlot()` function uses two horizon plots, with a shared data edge.

#### Polygons

## Empty plots

The uses of daily and yearly plots extend beyond linking data to earthly cycles. The following functions provide the basic coordinate system without any data plotted (these functions are used intenally by `dielPlot()` and `yearlyPlot()` to establish their coordinate system).

### `emptyDiel()`

```{r emptyDiel, fig.height=7, fig.width=9, fig.cap='An empty dielPlot().', fig.align='center', out.width='90%'}
emptyDiel()
```

### `emptyYearly()`

The `emptyYearly()` function takes an option `year` parameter that automatically adjusts the grid positions for leap years, although the visual effect is marginal.

```{r emptyYearly, fig.height=7, fig.width=9, fig.cap='An empty yearlyPlot().', fig.align='center', out.width='90%', cache=TRUE}
emptyYearly(year=2000)
```

## Adding data to the visualisation { #adding-to-cyclical }

### Periodic data: rings

The ring functions (`dielRing()`,...) plot ring segments on top of a base cyclical plot. These rings are useful for showing typical periods of activity for a species, or events that happen continuously for a specified period of time.

By defaults the limits for the rings are `1,2` for use with a _core_ type plot, but this can be changed by specifying the `limits` parameter to the ring function. Similarly, the plot legend may be removed with the paramater `legend=FALSE`.

#### `dielRings()`

```{r diel-plot-rings-1, fig.height=7, fig.width=10, fig.cap='A \'core\' diel plot with diel rings.', fig.align='center', out.width='90%', cache=TRUE}
names <- c("activity 1", "activity 2", "activity 3")
starts <- c("0600", "0900", "1500")
ends <- c("1200", "1700", "1900")
cols <- c("red", "green", "blue")

dielPlot("2022-08-08", lat=53, lon=0.1, limits=c(0,1))
dielRings(names, starts, ends, cols=cols)
```

```{r diel-plot-rings-2, fig.height=7, fig.width=10, fig.cap='A \'ring\' diel plot with diel rings.', fig.align='center', out.width='90%', cache=TRUE}
names <- c("activity 1", "activity 2", "activity 3")
starts <- c("0600", "0900", "1500")
ends <- c("1200", "1700", "1900")
cols <- c("red", "green", "blue")

dielPlot("2022-08-08", lat=53, lon=0.1, limits=c(1,2))
dielRings(names, starts, ends, cols=cols, limits=c(0,1))
```

### Periodic data: horizons

In this example we will plot the average monthly minimum and maximum temperatures for Lyme Regis, UK onto a `yearlyPlot()`. This example introduces two helper functions; `yearlyLabels()` and `circularise()`.

#### `yearlyLabels()` and `yearlyPositions`

These two functions are closely related, and used internally by `SonicScrewdriveR` to label a yearly plot.

```{r}
yearlyLabels()
```

The optional `pos` parameter when set to `TRUE` returns a vector of positions for the labels around the plot (in radians).

```{r}
yearlyPositions(year=2022)
```

The temperature data we have for Lyme Regis is monthly, however we would like to plot the values at the middle of the respective month. For this we add the parameter `format="mid-month"` to get the appropriate radial angles.

```{r}
yearlyPositions(year=2022, format="mid-months")
```

If we now plot this data we see that the output is not quite a complete ring or horizon data as we might have expected.

```{r lyme-temp-1, fig.height=7, fig.width=10, fig.cap='Horizon data plot without circularise().', fig.align='center', out.width='90%', cache=TRUE}
# Temperature data for Lyme Regis
t_min <- c(3, 2.7, 3.4, 5.2, 8.2, 11.2, 13.1, 13, 14.4, 9.2, 5.8, 3.8)
t_max <- c(7.9,8, 9.8, 12.4, 15.4, 18.2, 20.1, 19.5, 17.8, 14.4, 10.6, 8.4)

# Scale the data
sf <- max(t_max)
t_min <- t_min/sf
t_max <- t_max/sf

angles <- yearlyPositions(format="mid-months")

yearlyPlot(lat=50.7, lon=-2.9)
radialPolygon(angles, angles, t_min, t_max, col="orange")
```

#### `circularise()`

In order to join the horizons into a complete ring we can use the `circularise()` function on the angle and temperature vectors.

```{r lyme-temp-2, fig.height=7, fig.width=10, fig.cap='Horizon data plot with circularise().', fig.align='center', out.width='90%', cache=TRUE}
# Temperature data for Lyme Regis
t_min <- c(3, 2.7, 3.4, 5.2, 8.2, 11.2, 13.1, 13, 14.4, 9.2, 5.8, 3.8)
t_max <- c(7.9,8, 9.8, 12.4, 15.4, 18.2, 20.1, 19.5, 17.8, 14.4, 10.6, 8.4)

# Scale the data
sf <- max(t_max)
t_min <- t_min/sf
t_max <- t_max/sf

angles <- yearlyPositions(format="mid-months")

# Circularise
t_min <- circularise(t_min)
t_max <- circularise(t_max)
angles <- circularise(angles)

yearlyPlot(lat=50.7, lon=-2.9)
radialPolygon(angles, angles, t_min, t_max, col="orange")
```

### Helper functions

#### `dielFraction()`

The `dielFraction()` function is used to convert a POSIX time, or a time in `HHMM` string format, into a radial fraction. This function is called by `dielPlot()` and `dielRings()`. By default the output is multiplied by 2π to output a position round a circle.

```{r}
time <- Sys.time()
print(time)
frac <- dielFraction(time)
print(frac)
```
The raw fraction can be specified using the parameter `unit="fraction"`.

```{r}
time <- Sys.time()
print(time)
frac <- dielFraction(time, unit="fraction")
print(frac)
```

## Interactive Plots
These plots can be used to create Shiny apps

* [shiny-diel](https://shiny.ebaker.me.uk/shiny-diel/) is an example that shows diel plots for a number of locations, and can be animated using the play button under the date slider.
